Why does preregistration increase the persuasiveness of evidence? A Bayesian rationalization

### 2023-10-31 manuscript.Rmd

The last statement sounds like a truism but is directly related to Popper's seminal criterion of demarcation.
He stated that if it is impossible to prove that a hypothesis is false ($P(\neg{}E|\neg{}H) = 0$, theoretical risk is zero), it cannot be considered a scientific hypothesis [@popperLogicScientificDiscovery2002, p. 18].
We note these relations to underline that the Bayesian rationale we apply here is able to reconstruct many commonly held views on riskiness and epistemic value (but we of course differ from Popper on the central issue of induction in science).

@@

At the same time, approaching the evaluation of evidence using a Bayesian formalism is far from novel [e.g., @kuklaClinicalStatisticalTheory1990 @fiedlerWhatConstitutesStrong2017;@sprengerBayesianPhilosophyScience2019].
To our knowledge, it was not yet applied to the problem of preregistration.
However, @oberauerAddressingTheoryCrisis2019 made use of the formalism to model the relation between theory, hypothesis, and evidence.
In the context of this conceptualization, they discussed the usefulness of preregistration, though without applying the formalism there.
Most importantly, they are rather critical of the idea that preregistration has tangible benefits.
Instead, they prefer multiverse analyses but contend that those could be preregistered if one fancies it.
Their reasoning is based on two intuitions about what should *not* influence the evaluation of evidence: temporal order and the mental state of the originator.
In our opinion, they disregard the temporal order a bit too hastily, as it is a long-standing issue in Bayesian philosophy of science known as the "problem of old evidence" [@chiharaProblemsBayesianConfirmation1987].
However, we agree that not the temporal order is decisive but if the researchers incorporated the information into the hypothesis the evidence is supposed to confirm.
For the other, we argue that the mental state of the originator does matter.
Suppose there are $k = 1, 2, \ldots, K$ ways to analyze data.
Where each $k$ has a $P(E_k | \neg H) > 0$.
If they intend to try each way after another but happen to be "lucky" on the first try and stop, should we then apply $P(\neg E| \neg H) = P(\neg E_1 | \neg H)$ or $P(\neg E| \neg H) = P(\neg E_1 \land \ldots \land \neg E_k | \neg H)$?
We think the latter.
However, this "Defeatist" intuition is not universally warranted and depends on what we take $H$ to mean specifically [@kotzenMultipleStudiesEvidential2013].
Addressing, this problem might benefit from combining @oberauerAddressingTheoryCrisis2019's idea of updating on two nested levels (theory-hypothesis layered on top of hypothesis-evidence) with our approach to modelling uncertainty.

@@

We believe that our reasoning is quite similar to @hoflerMeansValuableExploration2022, who call for transparent exploration using preregistration.
We could be surer of our agreement, if they had formulated their arguments within a mathematical framework, which would also have helped to dissolve an apparent conflict in their definitions of confirmation, exploration, and transparency.
On the one hand, they define "The principle difference between confirmation and exploration is that confirmation adheres to an evidential norm for the test of a hypothesis to pass.", but then suggest that transparent exploration can be conducted using inferences tests as a filtering mechanism.
Their distinction between confirmation, intransparent and transparent exploration are otherwise just as well placed along the dimensions, theoretical risk and uncertainty about theoretical risk.

>>>>

The last statement sounds like a truism but is directly related to Popper's seminal criterion of demarcation.
He stated that if it is impossible to prove that a hypothesis is false ($P(\neg{}E|\neg{}H) = 0$, theoretical risk is zero), it cannot be considered a scientific hypothesis [@popperLogicScientificDiscovery2002, p. 18].
We note these relations to underline that the Bayesian rationale we apply here is able to reconstruct many commonly held views on riskiness and epistemic value (but we of course differ from Popper on the central role of induction in science).

@@

At the same time, approaching the evaluation of evidence using a Bayesian formalism is far from novel [e.g., @kuklaClinicalStatisticalTheory1990 @fiedlerWhatConstitutesStrong2017;@sprengerBayesianPhilosophyScience2019].
To our knowledge, it was not yet applied to the problem of preregistration.
However, @oberauerAddressingTheoryCrisis2019 made use of the formalism to model the relation between theory, hypothesis, and evidence.
In the context of this conceptualization, they discussed the usefulness of preregistration, though without applying the formalism there.
Most importantly, they are rather critical of the idea that preregistration has tangible benefits.
Instead, they prefer multiverse analyses but contend that those could be preregistered if one fancies it.
Their reasoning is based on two intuitions about what should *not* influence the evaluation of evidence: temporal order and the mental state of the originator.
In our opinion, they disregard the temporal order a bit too hastily, as it is a long-standing issue in Bayesian philosophy of science known as the "problem of old evidence" [@chiharaProblemsBayesianConfirmation1987].
However, we agree that not the temporal order is decisive but if the researchers incorporated the information into the hypothesis the evidence is supposed to confirm.
For the other, we argue that the mental state of the originator does matter.
Suppose there are $k = 1, 2, \ldots, K$ ways to analyze data, where each $k$ has a $P(E_k | \neg H) > 0$.
If they intend to try each way after another but happen to be "lucky" on the first try and stop, should we then apply $P(E| \neg H) = P(E_1 | \neg H)$ or $P(E| \neg H) = P(E_1 \lor \ldots \lor E_k | \neg H)$?
We think the latter.
However, this "Defeatist" intuition is not universally warranted and depends on what we take $H$ to mean specifically [@kotzenMultipleStudiesEvidential2013].
Addressing, this problem might benefit from combining @oberauerAddressingTheoryCrisis2019's idea of updating on two nested levels (theory-hypothesis layered on top of hypothesis-evidence) with our approach to modelling uncertainty.

@@

We believe that our reasoning is quite similar to @hoflerMeansValuableExploration2022, who call for transparent exploration using preregistration.
We could be more sure of our agreement, if they had formulated their arguments within a mathematical framework, which would also have helped to dissolve an apparent conflict in their definitions of confirmation, exploration, and transparency.
On the one hand, they define "The principle difference between confirmation and exploration is that confirmation adheres to an evidential norm for the test of a hypothesis to pass.", but then suggest that transparent exploration can be conducted using inferences tests as a filtering mechanism.
Their distinction between confirmation, intransparent and transparent exploration are otherwise just as well placed along the dimensions, theoretical risk and uncertainty about theoretical risk.
